Health Consequences and Economic Costs of Major Weather Events in the United States from 1950 to 2011.
========================================================

The following R markdown document analyzes and presents on major weather events that have occurred in the United States from 1950-2011. Bar charts are used to determine which types of events have caused the most casualties, crop damage and property damage. Casualties refers to a combination of injuries and fatalities that resulted from a specific event. Crop damage and property damage are presented in nominal US dollars. Only the top ten catastrophic events are displayed.

**Data Processing**

Lattice library is sourced for graphics:
```{r}
## Source 'lattice' library
library(lattice)
```
Data comes from online repository provided by Coursera instructors:
```{r}

## Set variables for download
url <- "http://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
datasetDirPath <- file.path("weather.csv.bz2")

## Downloading/Unzipping data *iff* data doesn't already exist
if(!file.exists(datasetDirPath)){ 
  download.file(url,datasetDirPath)
}

```

Weather data is read in as a CSV file first, where blanks cells are changed to 'NA'. A second, smaller data frame is created to work with. Column titles are also clarified.
```{r}

## Read in csv file
weather <- read.csv(bzfile(datasetDirPath), na.strings = c("NA", ""), stringsAsFactors = FALSE)

## Remove excess columns
weather <- data.frame(Event = weather$EVTYPE, Fatalities = weather$FATALITIES, Injuries = weather$INJURIES, PropertyDamage = weather$PROPDMG, propdmgexp = weather$PROPDMGEXP, CropDamage = weather$CROPDMG, cropdmgexp = weather$CROPDMGEXP, Remarks = weather$REMARKS)
```

For the analysis, any extraneous punctuation, or spaces and leading and trailing spaces are removed from the 'Event' field. 
```{r}
## Clean Event field
weather$Event <- gsub("[[:digit:][:punct:]])"," ", weather$Event)
weather$Event <- gsub("[[:space:]]+)"," ", weather$Event)
weather$Event <- gsub("^[[:space:]]+|[[:space:]]+$ )"," ", weather$Event)
```

Since the top ten events will be used, any duplicate or similarly labeled events that will appear in the top ten and change the resulting order will be combined.
```{r}
## Clean-up duplicates that will be in top ten.
weather$Event[grep("TSTM WIND|THUNDERSTORM WIND", weather$Event)] <- "TSTM WIND"
weather$Event[grep("WINTER STORM", weather$Event)] <- "BLIZZARD"
```

For the purposes of analyzing the health consequences, injuries and fatalities are combined into a more general 'casualties' column.

```{r}
## Calculate casualties from injury and fatality statistics
weather$Casualties <- weather$Injuries + weather$Fatalities
```

Next, the quantity of casualties by event type is determined, and ordered from most casualties to least. Rows without casualty data are dropped. Finally, the top ten events are subset.
```{r}
## Aggregate casualties by storm type
eventCasualties <- aggregate(Casualties ~ Event, weather, sum)
eventCasualties <- eventCasualties[eventCasualties$Casualties > 0,]
eventCasualties <- eventCasualties[with(eventCasualties, order(-Casualties, Event)), ]

## Pull Top Ten Events by Casualty
topEventsHealth <- head(eventCasualties, 10)
```

The economic consequences are divided into two categories: crop and property damage.

First, the monetary figures are converted to the same scale for the analysis
```{r}
## Create vector of amounts for multiplication
translate <- c(h = 100, k = 1000, m = 1e+06, b = 1e+09)
## Convert property damage to scale
weather$propdmgexp <- translate[tolower(weather$propdmgexp)]
weather$propdmgexp[is.na(weather$propdmgexp)] = 1

## Calculate dollar amount for property damage
weather$PropertyDamage <- weather$PropertyDamage * weather$propdmgexp

## Rinse and repeat for crop damage.
weather$cropdmgexp <- translate[tolower(weather$cropdmgexp)]
weather$cropdmgexp[is.na(weather$cropdmgexp)] = 1

weather$CropDamage <- weather$CropDamage * weather$cropdmgexp
```
Next, the crop and property damage are aggregated by event type.
```{r}
## Aggregate property and crop damage
eventPropDmg <- aggregate(PropertyDamage ~ Event, weather, sum)
eventPropDmg <- eventPropDmg[eventPropDmg$PropertyDamage > 0,]
eventPropDmg <- eventPropDmg[with(eventPropDmg, order(-PropertyDamage, Event)), ]

eventCropDmg <- aggregate(CropDamage ~ Event, weather, sum)
eventCropDmg <- eventCropDmg[eventCropDmg$CropDamage > 0,]
eventCropDmg <- eventCropDmg[with(eventCropDmg, order(-CropDamage, Event)), ]
```

Just as was done for casulaties, the top ten events by damage are taken.
```{r}
## Find top ten events for crop and property damage
topEventsProp <- head(eventPropDmg, 10)
topEventsCrop <- head(eventCropDmg, 10)
```

**Results**
The top ten events by casualty are presented below. Tornadic events are responsible for the most casualties.
```{r fig.width=7, fig.height=6}

## Graph
barchart(Casualties ~ Event, topEventsHealth, xlab="Event", ylab="Casualties")
```

The top ten events that cause the most crop damage are presented below. Droughts are a significant economic burden on crop production.
```{r fig.width=7, fig.height=6}
barchart(CropDamage ~ Event, topEventsCrop, xlab="Event", ylab="Crop Damage (US Dollars)")
```

The top ten events that cause the most property damage are presented below. Floods result in the most monetary damage to property.
```{r fig.width=7, fig.height=6}
barchart(PropertyDamage ~ Event, topEventsProp, xlab="Event", ylab="Property Damage (US Dollars)")
```

